### 2021-12-08 documents/sections/verbs.tex
trimming verbs

\subsubsection{\rc{eti} and \rc{a[p]} \qu{to be}}
\label{sec:be}
For a comprehensive comparative overview of these two roots, readers are refferd to \textcite[375--382]{gildea2018reconstructing}; they will not be discussed in detail here.
\rc{a[p]} is the original copula and can be reconstructed as already having various irregularities in \PC.
\rc{eti}, the other root, is reconstructed by \textcites{meira2009property}{gildea2018reconstructing} as originally meaning \qu{to dwell, live}, but serving as a copula already in \PC.\footnote{Such a stative, locative source is also suggested by the existence of \obj{it͡ʃi} \qu{to lie down} in \arara \parencite[196]{alves2017arara}.}
Various modern languages use reflexes of these two roots in a suppletive manner, conditioned by person and\slash{}or \gl{tam} value.
Both roots preserved \gl{1}\gl{s_a_} \rc{w-} in \PPek, \PWai, and \PTir (\crefrange{sec:pekodian}{sec:taranoan}).
\akuriyo \obj{a} was not affected by the extension of \obj{k-} \pcref{sec:akuriyo}, while \obj{eʔi} is not attested in a first-person form. 
\carijo innovated \obj{j-}, but only in the reflex of \rc{eti} \exref{mivida-12}; the \obj{a} root preserves \obj{w-} \pcref{sec:carijo}.
\yukpa introduced \obj{j-} to the reflexes of both \rc{a[p]} and \rc{eti}, which are preserved as encliticized auxiliaries in certain constructions \exref{yukcop}.

@@

\subsubsection{\rc{ka[ti]} \qu{to say}}
\label{sec:say}
Most reflexes of this verb are simply \obj{ka}, but a fleeting syllable \rc{ti} is reconstructed by \textcite{gildea2007greenberg}, best visible in the imperative forms of some languages \exref{kati}.

\ex<kati> \apalai\\
\obj{kaʃi-ko} \qu{say!}\\
\parencite[35]{koehn1986apalai}
\xe
%
\cref{tab:say} shows a comparison of the longest attested forms for each language.%
\footnote{Cognate segments in \crefrange{tab:say}{tab:bathe} were aligned automatically with LingPy \parencite{lingpy268}, for easier recognition of correspondences.}
%
\input{floats/say.tex}%
%
This verb was not affected by any of the extensions found in \PPek, \PWai, \PTir, \akuriyo, or \carijo (\crefrange{sec:pekodian}{sec:carijo}).
The first person form of its \yukpa reflex \obj{ka} is unattested.

@@

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}).
In \exref{hixsay.ordpl-35}, it does not occur, because \qu{they said} is preceded by direct speech.
Looking beyond \hixka, at least the \trio cognate shows the same pattern, albeit inconsistently so \parencite[267]{triocarlin2004}.

@@

The categorization of \qu{to say} as an intransitive verb is supported primarily by its person prefixes.
\kalina offers a minimal pair between transitive \obj{ka} \qu{to remove} and intransitive \obj{ka} \qu{to say}, \obj{sikai} \qu{I took it away} vs \obj{wɨkai} \qu{I said} \parencite[288, 45]{courtz2008carib}.
Similarly, \qu{to say} has a reflex of \gl{1}\gl{s_a_} \rc{w-} in Pekodian languages \pcref{sec:pekodian}, rather than \gl{1}>\gl{3} \obj{s-} (\bakairi) or \rc{ini-} (\PXin).
Additionally, languages which differentiate direct prefixes from \gl{s_a_} prefixes by the presence of \obj{i} \parencite[495]{meira2010origin} show \obj{ɨ} instead of \obj{i} for this verb, see \exref{sayintr.aku-154} as well as \textcites[294]{triomeira1999}[195]{wayanatavares2005}[288]{ikpengpacheco2001}[150]{alves2017arara}[168]{hoff1968carib} for the verb's inflection in such languages.
Finally, the \gl{s_a_} class marker \obj{w-} occurs on nominalizations in \kalina \exref{sayintr.kar-84}, and it is probably reflected as vowel length in the \trio \parencite[333]{triomeira1999} and \wayana \parencite[196]{wayanatavares2005} participles.

@@

In summary, this verb can be reconstructed as being intransitive based on its prefixes, but transitive based on some suffixes.
\hixka has lost the main intransitive criteria, making its reflex look more like a transitive verb.
It is not attested as being affected by any of the person marker extensions  in \cref{sec:extensions}.

\subsubsection{\rc{ɨtə[mə]} \qu{to go}}
\label{sec:go}
This verb is reconstructed by \textcite{gildea2007greenberg} as \rc{tə[mə]}, like \rc{ka[ti]} \qu{to say} with a fleeting second syllable.
It is true that many reflexes are clearly \obj{t}-initial, for example \hixka \obj{ntoje} \qu{he went} \parencite[27]{hixkaryanaderby1985}, \trio \obj{təkə} \qu{go!} \parencite[246]{triomeira1999}, or \wayana \obj{kuptəm} \qu{we went} \parencite[195]{wayanatavares2005}.
However, once one considers all forms of the various reflexes of this verb \pcref{tab:go}, an initial vowel \rc{ɨ} must clearly be reconstructed -- in contrast to unambiguously C-initial \rc{ka[ti]} \qu{to say}.%
\footnote{As indicated by the brackets in \cref{tab:go}, there are many languages where the initial vowel is only present in some forms.
Also, the prefix-verb boundary in many inflected forms like e.g.\ \trio \obj{wɨtənne} or \arara \obj{wɨdolɨ} \qu{I went} \parencites[43]{triomeira1999}[153]{alves2017arara} is ambiguous, since an epenthetic \obj{ɨ} breaks up potential CC clusters.
Still, when one considers unambiguous forms, the contrast with \rc{ka[ti]} becomes very clear.}
This verb was not affected by the any of the extensions discussed in \cref{sec:extensions}.

@@

This verb is reconstructed as \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all attested reflexes \pcref{tab:come} suggests a somewhat more complex story.
Crucially, the majority do not reflect the \rc{ət} part of their reconstruction, and many forms are ostensibly reflexes of \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ}, rather than \rc{epɨ}.
A unifying account of all these forms is made possible by reconstructing a \PC verb of the form \rc{(ət-)jəpɨ}, morphologically segmentable into a detransitivizing prefix and a root \rc{jəpɨ}.

Only the Pemongan languages and \kaxui point to an \rc{j}-initial root, and \kaxui \obj{johɨ} is very rare in contrast to more frequent \obj{o(o)hɨ}.
It only occurs in the third person form of the Progressive, meaning that the \obj{j} may be a reflex of third person \rc{i-}.
As \cref{tab:kaxprog} shows, regular \obj{o}-initial verbs have no third person prefix, while \obj{i-} occurs with C-initial \obj{to[mo]} \qu{to go}.
Even if \qu{to come} once shared this \obj{i-}, the circumstances are strongly in favor of a \obj{j}-initial \kaxui root, allowing the reconstruction of \PC \rc{jəpɨ}.
%
\input{floats/kaxprog}%
%\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
%\a<kax-82>
%\begingl
%\gla johɨ-rɨ//
%\glb \gl{3}.come-\gl{prog}//
%\glft \qu{S/he is coming.}//
%\endgl
%\a<kax-54>
%\begingl
%\gla o-w-ohɨ-rɨ//
%\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
%\glft \qu{You are coming.}//
%\endgl
%\a<kax-77>
%\begingl
%\gla i-nkɨ-rɨ//
%\glb \gl{3}-sleep-\gl{prog}//
%\glft \qu{S/he is sleeping.}//
%\endgl
%\a<kax-83>
%\begingl
%\gla Ø-osone-rɨ//
%\glb \gl{3}-dream-\gl{prog}//
%\glft \qu{S/he is dreaming.}//
%\endgl
%\xe
%
Most of the morphologically complex forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} do not show evidence for the segment \rc{j}, but the \obj{i} in the \akawaio form \obj{əsipɨ} is very likely a reflex of the sequence \rc{jə}.
This analysis is supported by the reflex \obj{ipɨ} from bare \rc{jəpɨ} in very closely related \macushi.

Turning to the many forms seemingly reflecting \rc{əpɨ} and \rc{epɨ} rather than \obj{jəpɨ}, it becomes evident are distributed widely in the family, sometimes even co-occurring in the same language.
A unifying account of these forms requires the root \rc{jəpɨ} to undergo two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
While both sound changes are found in other contexts throughout the family \parencite{meira2010origin}, they appear to have applied irregularly to this verb, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as a coalescence of \rc{e} and \rc{əe}.

@@

The interpretation of the \rc{ət} part as a detransitivizer is justifiable due to its form and its paradigmatically conditioned occurrence in some languages.
Although the combination of a detransitivizer and an intransitive verb seems semantically illogical, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, the \PC \gl{s_p_} verb \rc{wɨnɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
Also, \waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced the class membership of its reflexes in \kaxui, \kalina, \arara, \trio, and \panare \exref{pan-128}.

@@

Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
The scenario suggested here involves reflexes of the detransitivizer \rc{ət(e)-} being optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.
As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ}/\rc{jəpɨ} (\akuriyo) resisted the introduction of \PTir \rc{t-}.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come} and was thus not a potential target of innovative \rc{k-}.

>>>>

\subsubsection{\rc{eti} and \rc{a[p]} \qu{to be}}
\label{sec:be}
For a comprehensive comparative overview of these two roots, readers are referred to \textcite[375--382]{gildea2018reconstructing}; they will not be discussed in detail here.
\rc{a[p]} is the original copula and can be reconstructed as already having various irregularities in \PC.
\rc{eti} is reconstructed by \textcites{meira2009property}{gildea2018reconstructing} as originally meaning \qu{to dwell, live}, but serving as a copula already in \PC.\footnote{Such a stative, locative source is also suggested by the existence of \obj{it͡ʃi} \qu{to lie down} in \arara \parencite[196]{alves2017arara}.}
Various modern languages use reflexes of these two roots in a suppletive manner, conditioned by person and\slash{}or \gl{tam}.
Both roots preserved \gl{1}\gl{s_a_} \rc{w-} in \PPek, \PWai, and \PTir (\crefrange{sec:pekodian}{sec:taranoan}).
\akuriyo \obj{a} was not affected by the extension of \obj{k-} \pcref{sec:akuriyo}, while \obj{eʔi} is not attested in a first-person form. 
\carijo innovated \obj{j-}, but only in the reflex of \rc{eti} \exref{mivida-12}; the \obj{a} root preserves \obj{w-} \pcref{sec:carijo}.
\yukpa introduced \obj{j-} to the reflexes of both \rc{a[p]} and \rc{eti}, which are preserved as encliticized auxiliaries in certain constructions \exref{yukcop}.

@@

\subsubsection{\rc{ka[ti]} \qu{to say}}
\label{sec:say}
Most reflexes of this verb are simply \obj{ka}, but a fleeting syllable \rc{ti} is reconstructed by \textcite{gildea2007greenberg}, best visible in the imperative forms of some languages.% \exref{kati}.
%
%\ex<kati> \apalai\\
%\obj{kaʃi-ko} \qu{say!}\\
%\parencite[35]{koehn1986apalai}
%\xe
%
\cref{tab:say} shows a comparison of the longest attested forms for each language.%
\footnote{Cognate segments in \crefrange{tab:say}{tab:bathe} were aligned automatically with LingPy \parencite{lingpy268}, for easier recognition of correspondences.}
%
\input{floats/say.tex}%
%
This verb was not affected by any of the extensions found in \PPek, \PWai, \PTir, \akuriyo, or \carijo (\crefrange{sec:pekodian}{sec:carijo}).
The first person form of its \yukpa reflex \obj{ka} is unattested.

@@

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}).
It does not occur in \exref{hixsay.ordpl-35}, because \qu{they said} is preceded by direct speech.
Looking beyond \hixka, at least the \trio cognate shows the same pattern, albeit inconsistently so \parencite[267]{triocarlin2004}.

@@

The categorization of \qu{to say} as an intransitive verb is supported primarily by its person prefixes.
\kalina offers a minimal pair between transitive \obj{ka} \qu{to remove} and intransitive \obj{ka} \qu{to say}, \obj{sikai} \qu{I took it away} vs \obj{wɨkai} \qu{I said} \parencite[288, 45]{courtz2008carib}.
Similarly, \qu{to say} has a reflex of \gl{1}\gl{s_a_} \rc{w-} in Pekodian languages \pcref{sec:pekodian}, rather than \gl{1}>\gl{3} \obj{s-} (\bakairi) or \rc{ini-} (\PXin).
Additionally, languages which differentiate direct prefixes from \gl{s_a_} prefixes by the presence of \obj{i} \parencite[495]{meira2010origin} show \obj{ɨ} instead of \obj{i} for this verb, see \akuriyo in \exref{sayintr.aku-154}, as well as \textcites[294]{triomeira1999}[195]{wayanatavares2005}[288]{ikpengpacheco2001}[150]{alves2017arara}[168]{hoff1968carib} for cognate forms in other such languages.
Finally, the \gl{s_a_} class marker \obj{w-} occurs on nominalizations in \kalina \exref{sayintr.kar-84}, and it is probably reflected as vowel length in the \trio \parencite[333]{triomeira1999} and \wayana \parencite[196]{wayanatavares2005} participles.

@@

In summary, this verb can be reconstructed as being intransitive based on its (inflectional) prefixes, but transitive based on some (derivational) suffixes.
\hixka has lost the main intransitive criteria, making its reflex look more like a transitive verb.

\subsubsection{\rc{ɨtə[mə]} \qu{to go}}
\label{sec:go}
This verb is reconstructed by \textcite{gildea2007greenberg} as \rc{tə[mə]}, like \rc{ka[ti]} \qu{to say} with a fleeting second syllable.
It is true that many reflexes are clearly \obj{t}-initial, for example \hixka \obj{ntoje} \qu{he went} \parencite[27]{hixkaryanaderby1985}, \trio \obj{təkə} \qu{go!} \parencite[246]{triomeira1999}, or \wayana \obj{kuptəm} \qu{we went} \parencite[195]{wayanatavares2005}.
However, once one considers all forms of the various reflexes of this verb \pcref{tab:go}, an initial vowel \rc{ɨ} must clearly be reconstructed -- in contrast to unambiguously C-initial \rc{ka[ti]} \qu{to say}.%
\footnote{As indicated by the brackets in \cref{tab:go}, there are many languages where the initial vowel is only present in some forms.
Also, the prefix-verb boundary in many inflected forms like e.g.\ \trio \obj{wɨtənne} or \arara \obj{wɨdolɨ} \qu{I went} \parencites[43]{triomeira1999}[153]{alves2017arara} is ambiguous, since an epenthetic \obj{ɨ} breaks up potential CC clusters.
Still, when one considers unambiguous forms, the contrast with \rc{ka[ti]} becomes very clear.}
This verb was not affected by the any of the extensions in \cref{sec:extensions}.

@@

This verb is reconstructed as \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all attested reflexes \pcref{tab:come} points to a more complex story.
Crucially, the majority do not reflect the \rc{ət} part of their reconstruction, and many forms are ostensibly reflexes of \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ}, rather than \rc{epɨ}.
A unifying account of all these forms is achieved by reconstructing a \PC form \rc{(ət-)jəpɨ}, morphologically segmentable into a detransitivizing prefix and a root \rc{jəpɨ}.

Only the Pemongan languages and \kaxui point to an \rc{j}-initial root, and \kaxui \obj{johɨ} is very rare in contrast to more frequent \obj{o(o)hɨ}.
It only occurs in the third person form of the Progressive, meaning that the \obj{j} may be a reflex of third person \rc{i-}.
As \cref{tab:kaxprog} shows, regular \obj{o}-initial verbs have no third person prefix, while \obj{i-} occurs with C-initial \obj{to[mo]} \qu{to go}.
Even if \qu{to come} once shared this \obj{i-}, the circumstances are strongly in favor of a \obj{j}-initial \kaxui root, allowing the reconstruction of \PC \rc{jəpɨ}.
%
\input{floats/kaxprog}%
%\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
%\a<kax-82>
%\begingl
%\gla johɨ-rɨ//
%\glb \gl{3}.come-\gl{prog}//
%\glft \qu{S/he is coming.}//
%\endgl
%\a<kax-54>
%\begingl
%\gla o-w-ohɨ-rɨ//
%\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
%\glft \qu{You are coming.}//
%\endgl
%\a<kax-77>
%\begingl
%\gla i-nkɨ-rɨ//
%\glb \gl{3}-sleep-\gl{prog}//
%\glft \qu{S/he is sleeping.}//
%\endgl
%\a<kax-83>
%\begingl
%\gla Ø-osone-rɨ//
%\glb \gl{3}-dream-\gl{prog}//
%\glft \qu{S/he is dreaming.}//
%\endgl
%\xe
%
Most of the (non-Pemongan) morphologically complex forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} show no segmental trace of \rc{j}, but the \obj{i} in \akawaio \obj{əsipɨ} is likely a reflex of the sequence \rc{jə}.
This analysis is supported by the reflex \obj{ipɨ} from bare \rc{jəpɨ} in very closely related \macushi.

As for the many forms seemingly reflecting \rc{əpɨ} and \rc{epɨ} rather than \obj{jəpɨ}, they are distributed widely in the family, sometimes even co-occurring in the same language.
A unifying account requires the root \rc{jəpɨ} to undergo two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
While both sound changes are found in other contexts throughout the family \parencite{meira2010origin}, they appear to have applied irregularly to this verb, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as a coalescence of \rc{e} and \rc{əe}.

@@

The interpretation of the \rc{ət} part as a detransitivizer is based on its form and its paradigmatically conditioned occurrence in some languages.
Although the combination of a detransitivizer and an intransitive verb seems semantically illogical, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, the \PC \gl{s_p_} verb \rc{wɨnɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
Also, \waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced the class membership of its reflexes in \kaxui, \kalina, \arara, \trio, and \panare \exref{pan-128}.

@@

Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
The scenario suggested here involves reflexes of the detransitivizer \rc{ət(e)-} being optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.
As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ}/\rc{jəpɨ} (\akuriyo) resisted the introduction of \PTir \rc{t-}.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.
